Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-11-13 chapters/chapter06.tex
more meat in the indels

%PL: added this below, check for references
\subsection{Introducing covariates in the epoch model}

@@

%PL: incorporate the citations of this section

@@

Geophylogenies, or more generally geo-trees are spatially referenced phylogenies, for which taxa exist in a real space and can be displayed as cartographic elements within a map \citep{Kidd2010}.   
Possibly the first 
mapping of a phylogeny in a geographical context 
%diagram like that, depicting evolutionary tree over geographical map
is the famous \textit{Hypothetical Sketch of the Monophyletic Origin and the Extension of the 12 Races of Man from Lemuria over the Earth} shown in \cite{Haeckel1876} (see Figure~\ref{fig:lemuria}).
Ernst Haeckel, a XIX century naturalist entertained the idea that the absence of missing fossil links in human evolution at his time could be explained by existence of \textit{Lemuria}, a sunken island in the Indian Ocean.  
Although the hypothesis obviously did not hold, the phylogenetic visualization has received considerable attention.
%The phylogenetic visualisation has surpassed the, entirely discredited, work itself.

@@

% indel models
\subsection{Simulation of insertion deletion events}

In molecular evolution, not only substitution processes (Subsection \ref{sub:subst_models}), but also insertions and deletions (indels) provide a mechanism for generating genetic diversity.
%Whereas substitutions are point mutations which replace one nucleotide with another, indels proceed by either inserting or deleting nucleotides in a sequence, thus altering its length. 
Indel mutations are well known to have a significant impact on molecular evolutionary processes \citep{Fletcher2009}, and small viral genomes may be particularly sensitive to them.
%The extent of this impact will depends on the position of the event in the coding regions of the genome.
% FB: check if I'm not talking out of my ass here:
% PL: you are :), an indel that disrupts the reading frame makes the protein dysfunctional. If not, it still has a high probability of affecting the protein function
%When an integer multiple of a codon triplet is inserted or deleted it means that one or more amino acids are inserted or deleted leaving the protein function unchanged.
%In the case of one or two indel events the resulting gene usually codes for a different protein with different amino-acids.
Indels are not only of interest because of their evolutionary consequences, but they are also important to accurately position by alignment procedures.
%This makes insertion/deletion sequence simulation important for evaluating new models and alignment procedures.
Realistic indel simulation requires however specific stochastic models.
Standard modeling assumptions, such as those used in CTMC substitution models (see Subsection~\ref{sub:subst_models}), are not realistic for indel events. %PL:  point out which ones. Reversibility seems a major one
Future releases of $\pi$BUSS will aim at implementation stochastic processes for indel evolution. % that option, starting with the stochastic Dollo model \citep{LeQuesne1974}.
%PL: I deleted the stochastic doll model as it doesn't seem to be the first to think about. It is a specific acquisition-loss process that allows acquiring a trait once and then independently loosing that trait on multiple occasions (but never acquiring it anymore). A bit more on models specifically dealing with indel evolution and some references would suit better here. Quickly scan some papers to drop something on them -- you may get questions on this anyway.

>>>>

\subsection{Introducing covariates in the epoch model}

@@

@@

Geophylogenies, or more generally geo-trees are spatially referenced phylogenies, for which taxa exist in a real space and can be displayed as cartographic elements within a map \citep{Kidd2010}.   
Possibly the first 
mapping of a phylogeny in a geographical context 
%diagram like that, depicting evolutionary tree over geographical map
is the famous \textit{Hypothetical Sketch of the Monophyletic Origin and the Extension of the 12 Races of Man from Lemuria over the Earth} shown in \cite{Haeckel1876} (see Figure~\ref{fig:lemuria}).
Ernst Haeckel, a XIX century naturalist entertained the idea that the absence of missing fossil links in human evolution at his time could be explained by existence of \textit{Lemuria}, a sunken island in the Indian Ocean.  
% Although the hypothesis obviously did not hold, the phylogenetic visualization has received considerable attention.
%The phylogenetic visualisation has surpassed the, entirely discredited, work itself.
Although the hypothesis obviously did not hold, the phylogenetic visualization has received considerable attention and surpassed the discredited work.

@@

% indel models
\subsection{Simulation of insertion deletion events}
%The extent of this impact will depends on the position of the event in the coding regions of the genome.
% FB: check if I'm not talking out of my ass here:
% PL: you are :), an indel that disrupts the reading frame makes the protein dysfunctional. If not, it still has a high probability of affecting the protein function
%When an integer multiple of a codon triplet is inserted or deleted it means that one or more amino acids are inserted or deleted leaving the protein function unchanged.
%In the case of one or two indel events the resulting gene usually codes for a different protein with different amino-acids.

In molecular evolution, not only substitution processes (Subsection \ref{sub:subst_models}), but also insertions and deletions (indels) provide a mechanism for generating genetic diversity.
Indel mutations are well known to have a significant impact on molecular evolutionary processes \citep{Fletcher2009}, and small viral genomes may be particularly sensitive to them.
Indels are not only of interest because of their evolutionary consequences, but they are also important to accurately position by alignment procedures.
Realistic indel simulation requires however specific stochastic models.
Standard modeling assumptions, such as those used in CTMC substitution models (see Subsection~\ref{sub:subst_models}), are not realistic for indel events. %PL:  point out which ones. Reversibility seems a major one
Future releases of $\pi$BUSS will aim at implementation stochastic processes for indel evolution.
%PL: I deleted the stochastic doll model as it doesn't seem to be the first to think about. It is a specific acquisition-loss process that allows acquiring a trait once and then independently loosing that trait on multiple occasions (but never acquiring it anymore). A bit more on models specifically dealing with indel evolution and some references would suit better here. Quickly scan some papers to drop something on them -- you may get questions on this anyway.

Just like substitutions, indels are discrete event occuring in a real time.
By convention, following \cite{Thorne1991} the insertion-deletion events are typically presented as links separating character sequence bases (see Figure~\ref{fig:indels}).
Indels are characterized by two parameters.
The length represents the number of blocks inserted or deleted, where each block is an integer multiple of 1 for nucleotide or an integer multiple of 3 for codons space. 
Conversely the location is the position in the sequence where the blocks of characters are inserted or the position where the deletion begins.

\begin{figure}
\centering
$$\bullet\; C\;\star C\;\star\; A\;\star\; C\;\star\; T\;\star$$
\caption{
{ \footnotesize 
{\bf DNA sequence CCACT of length $L=5$, with $L+1$ links that separate the characters.} 
By convention to the right of each base there is a \textit{normal link} ($\star$); the leftmost base has an \textit{immortal link} to itself($\bullet$), ensuring that an insertion can occur at position $l=0$.
} % END: footnotesize
}
\label{fig:indels}
\end{figure}

\cite{dawg} models indel sizes \textit{i.a.} using a negative binomial distribution and a power law, truncated at some user specified, maximum indel size $M$.
\cite{Benner1993} notes that the power law (the Zipfian distribution, \cite{Gonnet1991}) follows the empirical gap lengths distribution.

If by $\lambda_{I}$ we denote the insertion rate at a location, then the waiting times until an insertion occurs are exponentially distributed with rate $\lambda_{I}(L+1)$.
Insertions are then drawn from the stationary distribution $\mathbf{\Pi}=\{\pi_{i},\ i\in\mathcal{E}\}$ of the underlying substitution model.

The procedure of \cite{dawg} simulates deletion by assuming that the simulated sequence, of length $L$ lies within a larger sequence of length $N$, where $L\ll N$.
The maximum allowed deletion spans for $M$ blocks, such that $M \ll N$.
This ensures that the deletions can happen at the end of the sequence.
A deletion of an arbitrary size $u$ happening in the larger sequence will result in deletions in the subsequence, if it begins at one of the L sites of the smaller sequence, or one of the $u-1$ sites preceding the subsequence.
Deletions are assumed to appear uniformly along the larger sequence, therefore the probability of a deletion of size $u$ in the larger sequence deleting some part of the smaller sequence is $\frac{u-1+L}{N}$.
If $\lambda_{D}$ is the rate of deletion per location then the waiting time for a deletion in a smaller sequence is exponentially distributed with rate $\lambda_{D}(u-1+L)$ \citep{dawg}.
